Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools

### 2024-03-25 paper.qmd
Revision EVT

---
title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
# subtitle: A case study of Ireland
bibliography:
  # cruse.bib
  - bibliography.bib
  - cruse.bib
  - references.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
execute: 
  echo: false
  message: false
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
  # html: default
  docx: default
# # # # Uncomment for formatted version
#   elsevier-pdf:
#     linestretch: 2
#     keep-tex: true
#     journal:
#       name: European Transport Research Review
#       formatting: preprint
#       model: 3p
#       cite-style: super
# # # #   # pdf: default

# # Uncomment the following line to produce the cover page
author:
  - name: Robin Lovelace
    email: r.lovelace@leeds.ac.uk
    corresponding: true
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Joey Talbot
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Eugeni Vidal-Tortosa
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Hussein Mahfouz
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Elaine Brick
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Peter Wright
    affiliations:
      - name: AECOM
        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
  - name: "Gary O'Toole"
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Dan Brennan
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
  - name: Suzanne Meade
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  This paper describes an approach for developing strategic cycle network planning tools. Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface. Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions. The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions. CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level. Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks. Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning. The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
date: last-modified
---

@@

<!-- Mobility kills. -->

Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

@@

Building cycle networks represents a relatively 'quick win' within the context of decarbonization [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient ebikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
Climate change mitigation is a major motivation for cycle network plans [@scappini2022].

@@

At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings of high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
7 in 10 trips in Ireland are by car [@national] .
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

@@

This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/) in the CRUSE Tool for Ireland, the landing page of which is illustrated in @fig-landing.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss limitations and possible future improvements to the approach, and the implications of the results for Ireland before making concluding remarks in @sec-conclusions.

# Tools for estimating cycling potential

Tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
These tools can be considered as a specific application of the concept of planning support systems (PSS) , which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.

Previous studies have applied the concept of PSS to specific modes, including walking [@bencekri2024], public transport [@barmentlo2012] and cycling [@bencekri2023].
While such prior work provides insight into the potential for approaches to transport planning that are both data-driven and participatory, the focus of this paper is on tools that focus attention on the *geographic distribution* of cycling potential and which answer the question "where to build".
Not all papers reviewed in this section classify themselves as, or even mention, PSS, but all of them can be considered as PSS.
This paper is informed by the thinking underlying PSS, including the idea that the resulting tools are more effective if they are modular and can inter-link with other tools and processes in the planning process: PSS can be conceived as a "toolbox with separate instruments that can work and talk to each other effectively (as in the open systems concept) and that will be selectively applied in varying configurations to support a particular phase in the planning process" [@geertman2002].
We share this conception of PSS, emphasizing the benefits of a focus on a particular mode (cycling) and a particular phase in the planning process (strategic network development).

Some of the most relevant existing tools are summarised below and presented in @tbl-tools.

@@

```{r}
#| label: tbl-tools
# For PDF:
# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
#   kableExtra::column_spec(1, width = "13em") |>
#   kableExtra::column_spec(2, width = "15em") |>
#   kableExtra::column_spec(3, width = "12em") |>
#   kableExtra::column_spec(4, width = "5em")
# For HTML:
knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
```

Area-based approaches were developed before network-based tools, such as a 2007 regression-based model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.
The approach also generated "forecasts for potential levels of bicycle use" with with increases in off-road routes, improvements of highway and, conversely, increases in car ownership and distance to work leading to decreases in cycling to work.

@@

A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymized Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people traveling to work and school at the Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to calculate a single route between the population weighted centroids of the zones associated with each OD pair.
This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
To tackle this issue we developed a new method for OD data disaggregation called 'jittering' [@lovelace2022].
The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.

@@

However, travel to school and work only constitute around only 30% of all trips in Ireland.
To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.

@@

We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimize journey time, accounting for traffic lights and surface type.
The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
The 'Balanced' route type is a compromize between the two, minimising journey time while avoiding the busiest roads.
Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.

@@

These statistics can be revealing: in Kildare it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
For the Fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.

@@

Illustration of the Quietest (left) and Fastest (right) route networks for Naas (top) and Newbridge (bottom).
:::

@@

The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall greenhouse gas emissions by 2030 as part of the pathway to net-zero emissions by 2050.
For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
In terms of car travel, the target is to "Increas\[e\] the proportion of kilometers driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometers driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.

To model this decrease in car driver km, cycling uptake increases in line with the Go Dutch scenario.
However, we only model shift from driving to cycling.
There is no shift from other modes of transport to cycling.

@@

Under the Go Dutch scenario cycling reaches levels equivalent to those found in the Netherlands, taking account of the effects of route hilliness (measured as mean gradient) and route distance.
This scenario uses the same model that was used PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].

@@

The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
For the main target audience, professional transport planners working at the county level, the tool provides a range of outputs, including estimates of cycling potential at the county and network level.
The provision of balanced (the default), quietest and fastest route networks enables planners to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure.
The tool also provides data downloads, enabling estimates of cycling potential on the network to be visualised and analysed with other tools such as QGIS, Python or R.

@@

The main result of the work presented in this paper is an open access web application and datasets on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
Like other national tools presented in @tbl-tools, a key feature of the results is that it provides a consistent and objective baseline for comparing cycling potentials in different areas.

The approach generates estimates of cycling potential on every road in the country and are therefore too extensive to present in their entirety in this paper.
Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' or barriers in the cycle network associated with a particular school, work place or other destination.

Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the resulting evidence, in @sec-quantiative.
We also present some qualitative results based on workshops and interviews with stakeholders in Ireland, in @sec-qualitative.

@@

## Selected quantitative results {#sec-quantiative}

@@

Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritize speed and directness over comfort and safety.
Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.

Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.

<!-- The results for the other -->
## Selected qualitative results {#sec-qualitative}

Thoughout the project, we undertook a series of workshops and interviews with stakeholders in Ireland to gather feedback on the approach and resulting tool.
The feedback was generally positive, with stakeholders commenting on the need for the tool, lack of data on cycling potential proving a barrier to decision-making and investment, and the potential for the tool.
At an end of project workshop in December 2023, stakeholders were asked to provide feedback on the tool and its potential use in practice.
Quotes and suggestions from these workshops, and a subsequent interview with a Senior Transport Planning using the evidence in their everyday work in Transport Infrastructure Ireland, and presented in this section (with quotes being attributed to named individuals where permission has been granted).

Early in the project, we presented the approach to practitioners in Kildare and Limmerick.
The feedback from the workshop in Kildare highlighted the need for the tool for practitioners at the county level, as highlighted in the following quote:

> "It's a missing piece of evidence that will help new projects get built" (Senior Engineer, Sustainable Transport & Traffic Management, Kildare County Council)

Practitioners in Kildare County Council that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications. They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network. Council officers expressed a preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes (users can switch between the on the 'Route types' page for each county although this functionality is not available in the landing page map due to other feedback requesting a simpler interface).

In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.

During the workshop with practitioners in Limerick, the estimates of Baseline and Go Dutch potential at the network levels were found to match well with local knowledge. The tool effectively highlighted the same sections on the transport network that experienced planners identified as priorities for investment in cycling, such as the South Circular Road. Limerick County Council expressed their interest in utilizing each network layer, particularly the quiet network, to support new cyclists who lack confidence to share space with motor traffic.
Specific suggestions from the workshop included the suggestion to align scenarios with national strategy (resulting in the Climate Action Plan scenario) and to provide data downloads provided as 'Shapefiles' for further analysis (now implemented although the download format is GeoPackage, an open standard for spatial data).

In a subsequent interview with a practitioner using the evidence generated by the CRUSE approach in their everyday work, we received the following feedback.
The outputs were a "great tool for getting started with project outline documents" and "ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
There were two suggestions from Declan which have not yet been implemented: the provision of data showing only "gold standard" POWSCAR data, and the need for training on the tool to ensure that all staff can use it effectively.
These, and other potential improvements, are discussed in the next section.

@@

The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.

@@

As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.

A limitation of the approach as implemented in the CRUSE tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].

>>>>

---
title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
# subtitle: A case study of Ireland
bibliography:
  # cruse.bib
  - bibliography.bib
  - cruse.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
execute: 
  echo: false
  message: false
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
  # html: default
  # docx: default
# # # # Uncomment for formatted version
  elsevier-pdf:
    linestretch: 2
    keep-tex: true
    journal:
      name: European Transport Research Review
      formatting: preprint
      model: 3p
      cite-style: super
# # # #   # pdf: default

# # Uncomment the following line to produce the cover page
author:
  - name: Robin Lovelace
    email: r.lovelace@leeds.ac.uk
    corresponding: true
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Joey Talbot
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Eugeni Vidal-Tortosa
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Hussein Mahfouz
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Elaine Brick
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Peter Wright
    affiliations:
      - name: AECOM
        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
  - name: "Gary O'Toole"
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Dan Brennan
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
  - name: Suzanne Meade
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  This paper describes an approach for developing strategic cycle network planning tools. Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface. Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions. The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions. CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level. Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks. Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning. The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
  - Transport Planning
  - Collaborative Planning
date: last-modified
---

@@

<!-- Mobility kills. -->
## Background

Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
Cars are a major and growing source of greenhouse gas (GHG) emissions [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018], and a cause of disease due to physical inactivity and air, noise and microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024]. <!-- Shouldn't these three sources share the brackets? --> Transport is responsible for 23% of GHG emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

@@

Building cycle networks represents a relatively 'quick win' within the context of decarbonization [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient ebikes --- which are seeing rapid uptake --- outperform electric cars, which are too heavy and expensive, for the majority of trips.
Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
Climate change mitigation is a major motivation for cycle network plans [@scappini2022].

@@

At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings of high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
Seven in ten trips in Ireland are by car [@national].
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

@@

## Aim and content

This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/).

In @sec-tools, we review existing current tools for estimating cycling potential.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss limitations and possible future improvements to the approach, and the implications of the results for Ireland before making concluding remarks in @sec-conclusions.

# Tools for estimating cycling potential {#sec-tools}

Tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
Some of the most relevant existing tools are summarised below and presented in @tbl-tools.

@@

```{r}
#| label: tbl-tools
knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
  kableExtra::column_spec(1, width = "13em") |>
  kableExtra::column_spec(2, width = "15em") |>
  kableExtra::column_spec(3, width = "12em") |>
  kableExtra::column_spec(4, width = "5em")
```

Area-based tools were developed first.
Parkin and colleagues presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.

@@

A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
This means that data with high levels of geographic resolution are necessary to estimate cyclical potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymized Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people traveling to work and school at the Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to calculate a single route between the population weighted centroids of the zones associated with each OD pair.
This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
To tackle this issue we developed a new method for OD data disaggregation called 'jittering'. <!--This reference should be removed to anonymise the paper. They will guess that Lovelace is involved, but saying we developed and then citing is as if we did not remove authors names from the paper :) --> The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.

@@

However, travel to school and work only constitute around only 30% of all trips in Ireland. <!-- Reference needed -->
To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips. <!-- This is already implemented, right?--> 
An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes. <!--This should be Table 2--->

@@

We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the fastest routes minimize journey time, accounting for traffic lights and surface type.
The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
The 'Balanced' route type is a compromize between the two, minimising journey time while avoiding the busiest roads.
Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.

@@

These statistics can be revealing: in Kildare it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
For the fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.

@@

Illustration of the quietest (left) and fastest (right) route networks for Naas (top) and Newbridge (bottom) in County Kildare.
:::

@@

The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall GHG emissions by 2030 as part of the pathway to net-zero emissions by 2050.
For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
In terms of car travel, the target is to "Increas\[e\] the proportion of kilometers driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometers driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.

To model this decrease in car km driven, cycling uptake increases in line with the Go Dutch scenario.
However, we only model shift from driving to cycling.
There is no shift from other modes of transport to cycling.

@@

Under the Go Dutch scenario cycling reaches levels equivalent to those found in the Netherlands, taking account of the effects of route hilliness (measured as mean gradient) and route distance.
This scenario uses the same model as the PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].

@@

The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
Its landing page is illustrated in @fig-landing).

Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
For the main target audience, professional transport planners working at the county level, the tool provides a range of outputs, including estimates of cycling potential at the county and network level.
The provision of balanced (the default), quietest and fastest route networks enables planners to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure.
The tool also provides data downloads, enabling estimates of cycling potential on the network to be visualised and analysed with other tools such as QGIS, Python or R.

@@

The main result of the work presented in this paper is an open access web application and datasets on current and potential future cycling levels in Ireland, with evidence provided at county and route segment levels.
Like the PCT on which CRUSE builds, a key feature of the results is that they are systematic and available for all counties, and most routes on which cycling is permitted, in Ireland.
The full set of results are therefore too extensive to present in their entirety in this paper.
Through the interactive web application, users can explore the data to generate the results that are most relevant to their needs, whether that is finding the cycling potential on a particular road or finding 'weak links' or barriers in the cycle network associated with a particular school, work place or other destination.

Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE Tool.

@@

@@

Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the fastest route network representing routes that prioritize speed and directness over comfort and safety.
Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.

Although Dublin City has made progress, there are still many parts of the fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.

<!-- The results for the other -->

@@

The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to fastest, balanced and quietest route networks.
This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.

@@

As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for tools such as CRUSE to grow.
However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.

A limitation of the approach as implemented in the CRUSE Tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
